# Introduction {#sec:introduction}

```{r draft-banner, echo=FALSE, results='asis'}
if (isTRUE(params$draft)) {
  label <- params$draft_label
  if (is.null(label) || !nzchar(label)) {
    label <- "DRAFT"
  }

  cat(paste0("**", label, "**\n\n"))
}
```

The New Salmon Escapement Database System (NuSEDS) stores salmon spawner survey records, spawner abundance estimates, and linkages between them for Pacific Region salmon populations [@openCanadaNuSEDS]. In NuSEDS, estimate Types 1--6 are used as a shorthand indicator of how interpretable an estimate is for downstream uses such as stock assessment and Wild Salmon Policy workflows.

In practice, Type assignment has varied across programs and areas because the historical table-only guidance does not fully specify how to treat incomplete coverage, uncertain methods, and other recurring real-world complications.

This update was prompted by inconsistent application across operational teams, which leads to inconsistent reporting in NuSEDS and increases the risk of downstream misinterpretation---especially near screening thresholds used in rapid status workflows [@pestal2023stateSalmon; @dfo2024rapidStatus].

## Purpose and scope

The purpose of this technical report is to provide forward, implementation-ready guidance for assigning NuSEDS estimate Types 1--6 in a way that is reproducible and interpretable to downstream data users.

This report:

- documents the decision logic for assigning Types 1--6, including standardized qualifier codes that record why an estimate was conservatively downgraded;
- clarifies evidence expectations for common operational complications;
- identifies a minimal set of metadata elements needed for interpretable Type assignment, with an emphasis on Stream Escapement Narratives (**SENs**) used to support NuSEDS uploads and review; and
- provides canonical roll-ups and mapping guidance for enumeration and estimate methods (including legacy-to-canonical term reconciliation) to support consistent forward use in NuSEDS and clearer interpretation of legacy method terms (Appendix C).

This report does **not** re-estimate escapement, evaluate biological status, or replace program-specific expert judgement.

## Hyatt (1997) intent and foundation

Hyatt (1997) developed a draft, hierarchical escapement classification key intended to be "unambiguous, easy to apply and reliable" for both field personnel and informed data users [@hyatt1997]. Hyatt noted that the key should be repeatable among users given the same information, useful both **before** surveys (planning) and **after** surveys (reporting/quality control), and applicable retrospectively to historic estimates stored in the regional database [@hyatt1997].

Hyatt framed classification around three linked dimensions: (i) properties of the survey and analytical methods used, (ii) statistical properties of the resulting estimate (units, accuracy, and precision), and (iii) the level of documentation supporting interpretation [@hyatt1997]. The NuSEDS estimate-type table is derived from this foundation. This report does **not** reproduce Hyatt's original draft key verbatim; instead, it operationalizes Hyatt's intent as a property-first decision sequence that can be applied consistently to contemporary workflows.

## Estimate type summary (Types 1--6)

Table 1 summarizes Types 1--6 for orientation. Types summarize interpretability given method properties, uncertainty evidence, and documentation; they do not guarantee unbiasedness.

```{r type-summary-table}
# NOTE: this table is for narrative orientation; the classification logic is defined in Appendix A.

type_summary <- data.frame(
  Type = c("1", "2", "3", "4", "5", "6"),
  Label = c(
    "Absolute abundance (census / near-census)",
    "Absolute abundance estimate (qualified)",
    "Relative abundance index (high resolution)",
    "Relative abundance index (medium resolution)",
    "Relative abundance / estimate (low evidence)",
    "Presence / not detected"
  ),
  Units = c(
    "Absolute abundance",
    "Absolute abundance (often with uncertainty)",
    "Relative units (index; within-series)",
    "Relative units (index; within-series)",
    "Relative units (uncertain)",
    "+ / -"
  ),
  Interpretation = c(
    "A direct count (or near-census) with high confidence that missed fish are negligible or well constrained.",
    "A standardized estimation method producing absolute abundance with qualified accuracy and, where available, quantified precision.",
    "A consistent, high-effort index that supports within-series comparisons; absolute abundance and uncertainty may not be fully quantified.",
    "A lower-effort index where comparability is more sensitive to timing, visibility, and coverage.",
    "A numeric estimate where method information, documentation, or key qualifiers are insufficient to support stronger interpretation.",
    "A non-numeric record indicating adults present or not detected with reliable species identification."
  ),
  check.names = FALSE
)

csas_kable(
  type_summary,
  escape = TRUE,
  bold_header = FALSE,
  font_size = 8,
  align = c("c", "l", "l", "l"),
  caption = "High-level summary of estimate Types 1--6 for orientation."
) |>
  csas_colspec(1, width = "1.0cm") |>
  csas_colspec(2, width = "4.0cm") |>
  csas_colspec(3, width = "3.4cm") |>
  csas_colspec(4, width = "6.6cm")
```

## Why an update is needed

In operational datasets, the same nominal estimate Type can be assigned to methodologically different estimates, and the historical table-only guidance does not encode several recurring qualifiers that affect interpretation. Key drivers include:

- **Modern method coverage:** hydroacoustic imaging sonar approaches (e.g., DIDSON/ARIS) are increasingly used, but are not always clearly represented in table-only guidance [@holmes2005didson].
- **Coverage/uptime and breach/bypass:** device outages, incomplete run-window coverage, and breach/bypass events can materially affect interpretability unless magnitude and correction methods are documented [@see2021stateSpace].
- **Timing, visibility, and effort standardization:** for survey-based indices, timing relative to run timing, observer efficiency, and visibility often govern comparability [@jones1998observer; @korman2002observerEfficiency; @holtCox2008visual].
- **Combined-method workflows:** multi-component estimates require explicit component recording to avoid concealing a weaker component [@parsonsSkalski2010escTechniques].
- **Downstream screening sensitivity:** estimate Types are used as data-quality filters in Wild Salmon Policy rapid status workflows; inconsistent assignment can create discontinuities near threshold cutoffs [@pestal2023stateSalmon; @dfo2024rapidStatus].

## How this guidance should be used when uploading to NuSEDS

This update is intended to support consistent Type assignment during data entry and review by making key qualifiers explicit.

In practice:

- **Record methods explicitly** (enumeration and estimation). If methods are unknown, the guidance assigns a conservative Type 5 and records a method-unknown qualifier.
- **Record key qualifiers** that affect interpretation (run-window coverage, uptime, breach/bypass context, timing/visibility constraints, and any infilling/interpolation method).
- **Provide documentation evidence** sufficient for independent interpretation (e.g., field logs, methods/QA notes). Where evidence is missing, the guidance applies conservative downgrades.
- **Report quantitative uncertainty when available** (e.g., CV or SE) to support qualified precision for Type 2 where appropriate (e.g., AUC and mark--recapture uncertainty reporting) [@english1992auc; @schwarz1993markRecapture; @hilborn1999periodic; @parken2003aucUncertainty].

## What this report provides

Relative to table-only interpretation, the updated guidance:

1. **Separates enumeration from estimation.** Enumeration methods (field data collection) are recorded distinctly from estimation methods (analysis and modelling), so eligibility rules and qualifiers are applied transparently.
2. **Defines method families.** The many enumeration and estimation methods observed in NuSEDS are grouped into a small set of method families to support consistent property-based checks (see the *Method families* section).
3. **Implements a property-first decision sequence.** In plain language: confirm record type (numeric vs presence/not detected), confirm methods are known, select a method family, apply method-family checks, and then apply final documentation/uncertainty gates (see the *Property-first decision key* section).
4. **Makes downgrade reasons explicit.** Conservative qualifiers (e.g., coverage/uptime, breach/bypass context, timing/visibility constraints, missing documentation, and quantified uncertainty when available) are recorded as explicit qualifier codes rather than as implicit judgement.
5. **Recommends minimal metadata for interpretability.** The guidance identifies a minimal set of metadata elements that must be captured for Type assignment to remain interpretable to downstream users (see the *Recommended metadata to record* section).
