# Classification framework and decision key

## Design principles

Hyatt (1997) intended estimate-type assignment to reflect (i) method properties, (ii) statistical properties, and (iii) documentation [@hyatt1997]. This report makes those dimensions explicit as decision gates, method-family checks, and final evidence requirements.

Practically, classification depends on **two method layers**:

- **Enumeration method** (field data collection)
- **Estimation method** (analysis/model that turns observations into an estimate)

```{r method-decomposition-figure, fig.cap="Conceptual decomposition of enumeration (field) methods and estimation (analysis) methods used to assign estimate Types and qualifiers.", fig.height=3.0, fig.width=6.0}
par(mar = c(0, 0, 0.5, 0))
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 6))

box <- function(x1, y1, x2, y2, label) {
  rect(x1, y1, x2, y2, col = "white", border = "black")
  text((x1 + x2) / 2, (y1 + y2) / 2, label, cex = 0.9)
}

box(0.5, 3.8, 4.2, 5.4, "Enumeration
(field method)
(e.g., fence,
visual, aerial)")
box(0.5, 1.0, 4.2, 2.6, "Evidence / qualifiers
(coverage, uptime,
breach/bypass,
timing, visibility,
documentation)")
box(5.8, 3.8, 9.5, 5.4, "Estimation
(analysis method)
(e.g., mark--recapture,
AUC, modelling)")
box(5.8, 1.0, 9.5, 2.6, "Type assignment
(1--6) +
qualifier codes")

arrows(4.2, 4.6, 5.8, 4.6, length = 0.08)
arrows(4.2, 1.8, 5.8, 1.8, length = 0.08)
arrows(2.4, 3.8, 2.4, 2.6, length = 0.08)
arrows(7.6, 3.8, 7.6, 2.6, length = 0.08)

text(5.0, 5.1, "decompose", cex = 0.8)
text(5.0, 2.3, "apply checks", cex = 0.8)
```

## Decision key design {#sec:decision-key}

The classification follows a property-first sequence:

1. **Data format gate**: non-numeric presence/not-detected is classified as Type 6.
2. **Method known gate**: if survey and analytical methods are not identified, the estimate is provisionally Type 5 and flagged as method-unknown.
3. **Method-family selection**: a primary method family scopes the applicable questions.
4. **Method-family checks**: coverage, effort, visibility, timing, and related criteria drive conservative downgrades.
5. **Final checks**: documentation and uncertainty evidence are evaluated; where evidence is missing, conservative downgrades apply.

This structure reduces subjective table interpretation by turning common qualifiers into explicit questions and recorded qualifier codes.

## Escapement Estimate Type Toolkit

An accompanying Escapement Estimate Type Toolkit (R Shiny application) applies the same decision logic described in this report and returns both (i) a **Final Type** (1--6) and (ii) qualifier codes explaining any conservative downgrades. The present report remains the authoritative guidance; the toolkit is a practical companion for user implementation.

Within the toolkit, three intermediate labels are used for workflow transparency: **Provisional Type** (the type supported by the method-family path before final gates), **Max Type** (the upper bound imposed by triggered downgrade rules), and **Candidate Type** (the type implied after combining the enumeration and estimation steps before final documentation/precision checks). The report is concerned with the **Final Type** reported after all qualifier enforcement.

## Guidance-to-implementation outlook

Implementation should align with the guidance and method-family checks in this report. Priorities are to:

- capture complete enumeration and estimation methods for each estimate;
- apply mandatory qualifiers where applicable; and
- track unresolved policy/specification items (e.g., mixed-method combination handling and ambiguity thresholds) through a controlled revision cycle.

## Handling common operational complications

The guidance treats several recurring complications as explicit qualifiers rather than implicit judgement calls.

### Missed Observations and infilling

Breach/bypass events at counting sites, and periods of missing observation due to outages or missed visits, can materially affect estimate quality and magnitude and correction methods must be documented. The updated guidance treats breach/bypass risk and incomplete coverage as explicit downgrade triggers and records whether defensible infilling/interpolation was required [@velezEspino2010atnarko; @holmes2005didson; @see2021stateSpace]. For example, run-window coverage below 95% of season days or device uptime below 95% are treated as explicit qualifiers, and any interpolation used to fill missing data should be recorded.

### Survey timing relative to run timing and bias

For survey-based indices, visit count alone is not sufficient: surveys must bracket the period when fish are present, and visibility constraints and observer effects can govern bias and comparability [@holtCox2008visual; @jones1998observer; @korman2002observerEfficiency; @hill1997aerialFrequency]. The updated guidance includes timing and visibility checks within method families where these factors are primary drivers of data quality and comparability.

### Combined-method estimates

Combined-method workflows (e.g., system-wide sonar with tributary visual apportionment, or fences supplemented during breach periods) should be documented. The updated guidance applies a conservative rule: where multiple components contribute to the final estimate, the assigned Type should not exceed that implied by the weakest component unless a documented integration method supports a higher classification [@parsonsSkalski2010escTechniques].

### Calibration and historical revisions

Where historical values have been recalibrated or revised, the updated guidance treats calibration as an analysis layer that should be captured in metadata (calibration source, diagnostics, and revision history) rather than as a new estimate Type. This supports interpretation of time-series values by downstream users, but this report does **not** prescribe automatic reclassification of existing records or a downstream assessment rule for calibrated series.

### Quantitative uncertainty

Hyatt (1997) distinguished higher-quality absolute-abundance estimates in part by qualified precision (variance evidence) [@hyatt1997]. Where available, reporting quantitative uncertainty (e.g., coefficient of variation [CV] and standard error [SE]) improves downstream interpretation and use. For example, AUC and peak-count methods have established approaches for incorporating uncertainty and evaluating sensitivity to survey timing and frequency [@english1992auc; @hilborn1999periodic; @hill1997aerialFrequency; @parken2003aucUncertainty; @millar2012auc]. The updated guidance includes a final precision/accuracy check and recommends capturing quantitative uncertainty when available.

## Recommended metadata elements for NuSEDS {#sec:metadata}

Most of the guidance can be expressed using existing NuSEDS fields, but several high-value metadata elements are needed to make estimate Types transparent and reproducible for downstream users.

The table below lists guidance-level metadata elements and the typical qualifier codes they support; detailed software rollout and database-change logistics are treated as implementation work outside the scope of this technical report.

```{r recommended-metadata-table}
recommended_meta <- data.frame(
  `Metadata element` = c(
    "Enumeration method (field)",
    "Estimation method (analysis)",
    "Primary method family (FS/V/A/S/T/R/P/M)",
    "Run-window coverage evidence",
    "Device/site uptime evidence",
    "Breach/bypass context",
    "Infilling/interpolation method",
    "Survey timing relative to run timing",
    "Visibility/conditions",
    "Documentation evidence",
    "Quantitative uncertainty (CV/SE)",
    "Combined-method components",
    "Calibration source + diagnostics",
    "Revision history (what changed/when)"
  ),
  `Implications for Interpretation` = c(
    "Scopes which property checks apply.",
    "Distinguishes analysis pathways (e.g., mark--recapture, AUC, modelling).",
    "Controls eligibility rules and the set of required qualifiers.",
    "Explains conservative downgrades due to missed start/end of run.",
    "Explains conservative downgrades due to outages/survey gaps.",
    "Explains conservative downgrades where missed fish are plausible.",
    "Distinguishes defensible correction from undocumented gap-filling.",
    "A core driver of bias and comparability for survey indices.",
    "Affects detectability and comparability (especially for visual methods).",
    "Determines whether a downstream user can independently interpret the estimate.",
    "Supports qualified precision for Type 2 and more transparent uncertainty.",
    "Prevents concealment of a weaker component within an overall estimate.",
    "Supports interpretation of recalibrated time series.",
    "Supports transparency when historical values have been revised."
  ),
  `Typical qualifier code(s)` = c(
    "METHOD_UNKNOWN (if missing)",
    "METHOD_UNKNOWN (if missing)",
    "METHOD_UNKNOWN (if missing)",
    "RUN_COVERAGE",
    "UPTIME",
    "BREACH_BYPASS",
    "INFILL_METHOD",
    "TIMING",
    "VISIBILITY / ENV_COND",
    "DOC",
    "PRECISION_ACCURACY",
    "Component-level qualifiers + integration rule",
    "Calibration diagnostics / revision metadata",
    "Revision metadata / audit trail"
  ),
  `Priority for upload` = c(
    "Required",
    "Required",
    "Required",
    "Recommended",
    "Recommended",
    "Recommended",
    "Recommended",
    "Recommended",
    "Recommended",
    "Required",
    "Recommended",
    "Recommended",
    "Recommended",
    "Recommended"
  ),
  check.names = FALSE
)

csas_kable(
  recommended_meta,
  escape = TRUE,
  bold_header = FALSE,
  font_size = 8,
  align = c("l", "l", "l", "c"),
  caption = "Recommended metadata elements for NuSEDS uploads to support transparent Type 1--6 assignment and interpretation. Some of these are already present in NuSEDS."
) |>
  csas_colspec(1, width = "3.7cm") |>
  csas_colspec(2, width = "7.0cm") |>
  csas_colspec(3, width = "3.0cm") |>
  csas_colspec(4, width = "2.3cm")
```

## Method families {#sec:method-families}

Families are ordered by best attainable Type to foreground the upper bound implied by the guidance; short codes are retained for toolkit use.

```{r method-family-table}
method_families <- data.frame(
  Code = c("FS", "V", "S", "T", "M", "A", "R", "P"),
  `Method family` = c(
    "Fixed site census / near-census (manual or validated counter tally)",
    "Visual ground or snorkel count",
    "Hydroacoustic sonar count (modelled)",
    "Trap model (non-spanning)",
    "Mark-recapture program",
    "Aerial survey count",
    "Redd survey",
    "Electrofishing CPUE index"
  ),
  `Best attainable type` = c("1", "2", "2", "2", "2", "3", "3", "3"),
  check.names = FALSE
)

csas_kable(
  method_families,
  escape = TRUE,
  bold_header = FALSE,
  font_size = 9,
  align = c("c", "l", "c"),
  caption = "Method families encoded in the property-first guidance and the best attainable Type before conservative downgrades."
) |>
  csas_colspec(1, width = "2.4cm") |>
  csas_colspec(2, width = "9.6cm") |>
  csas_colspec(3, width = "2.5cm")
```

Type 1 retains an interpretive meaning---**absolute abundance from a census / near-census program**---rather than serving as a raw method label. In practice, only constrained fixed-site tallies can default to that upper bound under the current guidance.

Estimation method labels such as AUC and peak-count do not replace the underlying survey family. Ground/snorkel series analysed with AUC can still support Type 2 under the `V` family, whereas aerial series analysed with AUC or peak-count remain capped by the `A` family ceiling (Type 3) unless a future validated exception is explicitly adopted.

Default redd-count pathways are conservatively capped at Type 3 in this update. Any higher-confidence redd-derived absolute-abundance estimate should be treated as an explicit calibrated/model-based exception rather than as the default redd survey interpretation.

## DFO Salmon Data Standards and Controlled Vocabularies linkages for methods and downgrade criteria

To support consistent interpretation (including cross-program interoperability), these NuSEDS elements can be linked to the DFO Salmon Data Standards and Controlled Vocabularies (DFO Salmon Ontology) term browser and concept schemes:

- <https://w3id.org/gcdfo/salmon>

```{r ontology-linkage-schemes-table}
ontology_schemes <- data.frame(
  `NuSEDS element` = c(
    "ENUMERATION_METHODS",
    "ESTIMATE_METHOD",
    "DOWNGRADE_CRITERIA"
  ),
  `Ontology scheme term` = c(
    "gcdfo:EnumerationMethodScheme",
    "gcdfo:EstimateMethodScheme",
    "gcdfo:DowngradeCriteriaScheme"
  ),
  `Ontology IRI` = c(
    "https://w3id.org/gcdfo/salmon#EnumerationMethodScheme",
    "https://w3id.org/gcdfo/salmon#EstimateMethodScheme",
    "https://w3id.org/gcdfo/salmon#DowngradeCriteriaScheme"
  ),
  `WIDOCO term page` = c(
    "https://dfo-pacific-science.github.io/dfo-salmon-ontology/#EnumerationMethodScheme",
    "https://dfo-pacific-science.github.io/dfo-salmon-ontology/#EstimateMethodScheme",
    "https://dfo-pacific-science.github.io/dfo-salmon-ontology/#DowngradeCriteriaScheme"
  ),
  check.names = FALSE
)

csas_kable(
  ontology_schemes,
  escape = TRUE,
  bold_header = FALSE,
  font_size = 8,
  align = c("l", "l", "l", "l"),
  caption = "Ontology concept-scheme linkages for NuSEDS enumeration methods, estimate methods, and downgrade criteria."
) |>
  csas_colspec(1, width = "3.2cm") |>
  csas_colspec(2, width = "3.8cm") |>
  csas_colspec(3, width = "4.2cm") |>
  csas_colspec(4, width = "4.2cm")
```

Detailed term-level crosswalks are provided in Appendix B (downgrade criteria) and Appendix C (enumeration and estimate methods). Appendix C also includes compact Sankey-style summaries that illustrate how legacy NuSEDS method terms align with the standardized parent categories used by the updated guidance. Equivalent lookup helpers are also exposed in the `metasalmon` package (`nuseds_enumeration_method_crosswalk` and `nuseds_estimate_method_crosswalk`) [@metasalmonPackage].

Legacy values are not dropped; where a legacy term does not cleanly align to a current method family, it can be represented as inherited contextual guidance via ontology metadata (for example via `rdfs:comment`) while remaining traceable in source records.  

## NuSEDS data dictionary alignment (selected fields)

The NuSEDS data dictionary defines the database fields used to store enumeration methods, estimation methods, estimate classification (Types 1--6), and supporting metadata (e.g., inspections/effort and timing fields) [@openCanadaNuSEDSDataDictionary]. The updated guidance is designed to be expressible using existing NuSEDS fields where possible, and to clearly identify gaps where additional metadata would improve interpretation and reproducibility.

```{r nuseds-fields-table}
dd_path <- file.path("docs", "context", "Data_Dictionary_NuSEDS_EN.csv")

wanted_fields <- c(
  "ESTIMATE_CLASSIFICATION",
  "ENUMERATION_METHODS",
  "ESTIMATE_METHOD",
  "NO_INSPECTIONS_USED",
  "INDEX_YN",
  "ACCURACY",
  "PRECISION",
  "RELIABILITY",
  "ESTIMATE_STAGE"
)

field_roles <- data.frame(
  `Field Name` = wanted_fields,
  `Role in updated guidance` = c(
    "Stores Type 1--6 estimate classification (including legacy non-Type labels where present)",
    "Primary field method (enumeration) used to scope method-family checks",
    "Primary analysis method (estimation), including combined or calibrated special cases",
    "Supports effort/visit thresholds (VISITS and related downgrades)",
    "Flags index (partial-coverage) estimates (relative-abundance context)",
    "Legacy qualitative uncertainty field; not a substitute for quantified metadata",
    "Legacy qualitative uncertainty field; not a substitute for quantified metadata",
    "Legacy/import field (historical); not consistently present",
    "QA/workflow stage (preliminary/near final/final); not a type determinant"
  ),
  check.names = FALSE
)

if (file.exists(dd_path)) {
  dd <- read.csv(
    dd_path,
    stringsAsFactors = FALSE,
    check.names = FALSE,
    fileEncoding = "latin1"
  )

  dd_small <- dd[dd[["Field Name"]] %in% wanted_fields, c("Field Name"), drop = FALSE]
  dd_small <- merge(dd_small, field_roles, by = "Field Name", all.x = TRUE)
  dd_small <- dd_small[match(wanted_fields, dd_small[["Field Name"]]), ]
} else {
  dd_small <- data.frame(
    `Field Name` = wanted_fields,
    check.names = FALSE
  )
  dd_small <- merge(dd_small, field_roles, by = "Field Name", all.x = TRUE)
  dd_small <- dd_small[match(wanted_fields, dd_small[["Field Name"]]), ]
}

rownames(dd_small) <- NULL

csas_kable(
  dd_small,
  escape = TRUE,
  bold_header = FALSE,
  font_size = 8,
  align = c("l", "l"),
  caption = "NuSEDS fields relevant to estimate-type classification and how they relate to the updated guidance (see the NuSEDS data dictionary in docs/context/)."
) |>
  csas_colspec(1, width = "3.5cm") |>
  csas_colspec(2, width = "10.5cm")
```
