# Methods

## Design principles (Hyatt-aligned)

Hyatt (1997) intended estimate-type assignment to reflect (i) method properties, (ii) statistical properties, and (iii) documentation [@hyatt1997]. The updated guidance encodes these dimensions explicitly as decision gates, method-family checks, and final evidence requirements.

A practical implication of this structure is that classification depends on **two distinct method layers**:

- **Enumeration method** (field data collection), and
- **Estimation method** (analysis/model that turns observations into an estimate).

```{r method-decomposition-figure, fig.cap="Conceptual decomposition of enumeration (field) methods and estimation (analysis) methods used to assign estimate Types and qualifiers.", fig.height=3.0, fig.width=6.8}
par(mar = c(0, 0, 0.5, 0))
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 6))

box <- function(x1, y1, x2, y2, label) {
  rect(x1, y1, x2, y2, col = "white", border = "black")
  text((x1 + x2) / 2, (y1 + y2) / 2, label, cex = 0.9)
}

box(0.5, 3.8, 4.2, 5.4, "Enumeration
(field method)
(e.g., fence,
visual, aerial)")
box(0.5, 1.0, 4.2, 2.6, "Evidence / qualifiers
(coverage, uptime,
breach/bypass,
timing, visibility,
documentation)")
box(5.8, 3.8, 9.5, 5.4, "Estimation
(analysis method)
(e.g., mark--recapture,
AUC, modelling)")
box(5.8, 1.0, 9.5, 2.6, "Type assignment
(1--6) +
qualifier codes")

arrows(4.2, 4.6, 5.8, 4.6, length = 0.08)
arrows(4.2, 1.8, 5.8, 1.8, length = 0.08)
arrows(2.4, 3.8, 2.4, 2.6, length = 0.08)
arrows(7.6, 3.8, 7.6, 2.6, length = 0.08)

text(5.0, 5.1, "decompose", cex = 0.8)
text(5.0, 2.3, "apply checks", cex = 0.8)
```

## Decision key design {#sec:decision-key}

The classification follows a property-first sequence:

1. **Data format gate**: non-numeric presence/not-detected is classified as Type 6.
2. **Method known gate**: if survey and analytical methods are not identified, the estimate is provisionally Type 5 and flagged as method-unknown.
3. **Method family selection**: a primary method family scopes the applicable questions.
4. **Method-family checks**: coverage, effort, visibility, timing, and related criteria drive conservative downgrades.
5. **Final checks**: documentation and uncertainty evidence are evaluated; where evidence is missing, conservative downgrades apply.

This structure is intended to reduce subjective table interpretation by turning common qualifiers into explicit questions and recorded qualifier codes.

## Companion implementation (informative)

A companion internal implementation (R Shiny application) applies the same decision logic described here and returns both (i) a final estimate Type (1--6) and (ii) explicit qualifier codes explaining any conservative downgrades. The software is a practical companion tool; it is **not** the primary product of this report.

## Guidance-to-implementation outlook

Where possible, implementation should align with the guidance and the method-family checks in this report. The priority implementation tasks are to:

- capture complete enumeration and estimation methods for each estimate,
- apply all mandatory qualifiers where applicable,
- monitor unresolved policy/specification items (for example mixed-method combination handling and ambiguity thresholds) and close them through a controlled revision cycle.

## Handling common operational complications

The updated guidance treats several recurring complications as explicit qualifiers rather than implicit judgement calls.

### Breaches/bypass and infilling

Breach/bypass events at counting sites, and periods of missing observation due to outages or missed visits, can materially affect interpretability unless the magnitude and correction method are documented. The updated guidance treats bypass/breach risk and incomplete coverage as explicit downgrade triggers and records whether defensible infilling/interpolation was required [@velezEspino2010atnarko; @holmes2005didson; @see2021stateSpace]. For implementation, the guidance recommends recording explicit coverage/uptime qualifiers (e.g., run-window coverage below 95% of season days; device uptime below 95%) and whether interpolation was required for missing data.

### Survey timing relative to run timing

For survey-based indices, visit count alone is not sufficient: surveys must bracket the period when fish are present, and visibility constraints and observer effects can govern bias and comparability [@holtCox2008visual; @jones1998observer; @korman2002observerEfficiency; @hill1997aerialFrequency]. The updated guidance includes timing and visibility checks within method families where these factors are primary drivers of interpretability.

### Combined-method estimates

Combined-method workflows (e.g., system-wide sonar with tributary visual apportionment, or fences supplemented during breach periods) should be recorded as explicit components. The updated guidance applies a conservative rule: where multiple components contribute to the final estimate, the assigned Type should not exceed that implied by the weakest component unless a documented integration method supports a higher classification [@parsonsSkalski2010escTechniques].

### Calibration and historical revisions

Where historical values have been recalibrated or revised, the updated guidance treats calibration as an analysis layer that should be captured in metadata (calibration source, diagnostics, and revision history) rather than as a new estimate Type. This supports interpretation of time-series values by downstream users.

### Quantitative uncertainty (CV/SE)

Hyatt (1997) distinguished higher-quality absolute-abundance estimates in part by qualified precision (variance evidence) [@hyatt1997]. Where available, reporting quantitative uncertainty (e.g., CV or SE) improves interpretability for downstream use. For example, AUC and peak-count approaches have established approaches for incorporating uncertainty and evaluating sensitivity to survey timing and frequency [@english1992auc; @hilborn1999periodic; @hill1997aerialFrequency; @parken2003aucUncertainty; @millar2012auc]. The updated guidance includes a final precision/accuracy check and recommends capturing quantitative uncertainty when available.

## Recommended metadata elements for NuSEDS {#sec:metadata}

The updated guidance can be expressed using existing NuSEDS fields in many cases, but several high-value metadata elements are needed to make estimate Types interpretable and reproducible for downstream users.

**Note:** The table below lists guidance-level metadata elements and the typical qualifier codes they support. Detailed software rollout and database-change logistics are treated as companion implementation work, not as a primary product of this technical report.

```{r recommended-metadata-table}
recommended_meta <- data.frame(
  `Metadata element` = c(
    "Enumeration method (field)",
    "Estimation method (analysis)",
    "Primary method family (FS/V/A/S/T/R/P/M)",
    "Run-window coverage evidence",
    "Device/site uptime evidence",
    "Breach/bypass context",
    "Infilling/interpolation method",
    "Survey timing relative to run timing",
    "Visibility/conditions",
    "Documentation evidence",
    "Quantitative uncertainty (CV/SE)",
    "Combined-method components",
    "Calibration source + diagnostics",
    "Revision history (what changed/when)"
  ),
  `Why it matters for interpretation` = c(
    "Scopes which property checks apply.",
    "Distinguishes analysis pathways (e.g., mark--recapture, AUC, modelling).",
    "Controls eligibility rules and the set of required qualifiers.",
    "Explains conservative downgrades due to missed start/end of run.",
    "Explains conservative downgrades due to outages/unobserved intervals.",
    "Explains conservative downgrades where missed fish are plausible.",
    "Distinguishes defensible correction from undocumented gap-filling.",
    "A core driver of bias and comparability for survey indices.",
    "Affects detectability and comparability (especially for visual methods).",
    "Determines whether a downstream user can independently interpret the estimate.",
    "Supports qualified precision for Type 2 and more transparent uncertainty.",
    "Prevents concealment of a weaker component within an overall estimate.",
    "Supports interpretation of recalibrated time series.",
    "Supports transparency when historical values have been revised."
  ),
  `Typical qualifier code(s)` = c(
    "METHOD_UNKNOWN (if missing)",
    "METHOD_UNKNOWN (if missing)",
    "METHOD_UNKNOWN (if missing)",
    "RUN_COVERAGE",
    "UPTIME",
    "BREACH_BYPASS",
    "INFILL_METHOD",
    "TIMING",
    "VISIBILITY / ENV_COND",
    "DOC",
    "PRECISION_ACCURACY",
    "(varies; conservative component rule)",
    "(varies)",
    "(varies)"
  ),
  `Priority for upload` = c(
    "Required",
    "Required",
    "Required",
    "Recommended",
    "Recommended",
    "Recommended",
    "Recommended",
    "Recommended",
    "Recommended",
    "Required",
    "Recommended",
    "Recommended",
    "Recommended",
    "Recommended"
  ),
  check.names = FALSE
)

csas_kable(
  recommended_meta,
  escape = TRUE,
  bold_header = FALSE,
  font_size = 8,
  align = c("l", "l", "l", "c"),
  caption = "Recommended metadata elements for NuSEDS uploads to support transparent Type 1--6 assignment and interpretation."
) |>
  csas_colspec(1, width = "3.7cm") |>
  csas_colspec(2, width = "7.0cm") |>
  csas_colspec(3, width = "3.0cm") |>
  csas_colspec(4, width = "2.3cm")
```

## Method families {#sec:method-families}

```{r method-family-table}
method_families <- data.frame(
  Code = c("FS", "V", "A", "S", "T", "R", "P", "M"),
  `Method family` = c(
    "Fixed site census (manual or electronic)",
    "Visual ground or snorkel count",
    "Aerial survey count",
    "Hydroacoustic sonar count (modelled)",
    "Trap model (non-spanning)",
    "Redd survey",
    "Electrofishing CPUE index",
    "Mark-recapture program"
  ),
  `Best attainable type` = c("1", "2", "3", "2", "2", "2", "3", "2"),
  check.names = FALSE
)

csas_kable(
  method_families,
  escape = TRUE,
  bold_header = FALSE,
  font_size = 9,
  align = c("c", "l", "c"),
  caption = "Method families encoded in the property-first guidance and the best attainable Type before conservative downgrades."
) |>
  csas_colspec(1, width = "1.2cm") |>
  csas_colspec(2, width = "10.8cm") |>
  csas_colspec(3, width = "2.5cm")
```

## How this will proceed

Near-term implementation will proceed in a narrow order so the guidance can be applied consistently before extending scope:

- prioritize SEN-only fields needed to operationalize transparent Type assignment and conservative downgrades,
- finalize controlled lookup tables for those fields so downstream implementation can remain interoperable,
- defer broader SIL/CU enhancements to a later governance cycle.

In practical terms, this keeps this technical report focused on interpretability and leaves software rollout details to companion implementation tracks.

## Ontology linkages for methods and downgrade criteria

To support consistent interpretation (including cross-program interoperability), these NuSEDS elements can be linked to published DFO Salmon Ontology concept schemes and term pages:

- ontology entrypoint: <https://w3id.org/gcdfo/salmon>
- human-readable term documentation: <https://dfo-pacific-science.github.io/dfo-salmon-ontology/>

```{r ontology-linkage-schemes-table}
ontology_schemes <- data.frame(
  `NuSEDS element` = c(
    "ENUMERATION_METHODS",
    "ESTIMATE_METHOD",
    "DOWNGRADE_CRITERIA"
  ),
  `Ontology scheme term` = c(
    "gcdfo:EnumerationMethodScheme",
    "gcdfo:EstimateMethodScheme",
    "gcdfo:DowngradeCriteriaScheme"
  ),
  `Ontology IRI` = c(
    "https://w3id.org/gcdfo/salmon#EnumerationMethodScheme",
    "https://w3id.org/gcdfo/salmon#EstimateMethodScheme",
    "https://w3id.org/gcdfo/salmon#DowngradeCriteriaScheme"
  ),
  `WIDOCO term page` = c(
    "https://dfo-pacific-science.github.io/dfo-salmon-ontology/#EnumerationMethodScheme",
    "https://dfo-pacific-science.github.io/dfo-salmon-ontology/#EstimateMethodScheme",
    "https://dfo-pacific-science.github.io/dfo-salmon-ontology/#DowngradeCriteriaScheme"
  ),
  check.names = FALSE
)

csas_kable(
  ontology_schemes,
  escape = TRUE,
  bold_header = FALSE,
  font_size = 8,
  align = c("l", "l", "l", "l"),
  caption = "Ontology concept-scheme linkages for NuSEDS enumeration methods, estimate methods, and downgrade criteria."
) |>
  csas_colspec(1, width = "3.2cm") |>
  csas_colspec(2, width = "3.8cm") |>
  csas_colspec(3, width = "4.2cm") |>
  csas_colspec(4, width = "4.2cm")
```

Detailed term-level crosswalks are provided in Appendix B (downgrade criteria) and Appendix D (enumeration and estimate methods).

## NuSEDS data dictionary alignment

The NuSEDS data dictionary defines the database fields used to store enumeration methods, estimation methods, estimate classification (Types 1--6), and supporting metadata (e.g., inspections/effort and timing fields) [@openCanadaNuSEDSDataDictionary]. The updated guidance is designed to be expressible using existing NuSEDS fields where possible, and to clearly identify gaps where additional metadata would improve interpretation and reproducibility.

```{r nuseds-fields-table}
dd_path <- file.path("docs", "context", "Data_Dictionary_NuSEDS_EN.csv")

wanted_fields <- c(
  "ESTIMATE_CLASSIFICATION",
  "ENUMERATION_METHODS",
  "ESTIMATE_METHOD",
  "NO_INSPECTIONS_USED",
  "INDEX_YN",
  "ACCURACY",
  "PRECISION",
  "RELIABILITY",
  "ESTIMATE_STAGE"
)

field_roles <- data.frame(
  `Field Name` = wanted_fields,
  `Role in updated guidance` = c(
    "Stores Type 1--6 estimate classification (including legacy non-Type labels where present)",
    "Primary field method (enumeration) used to scope method-family checks",
    "Primary analysis method (estimation), including combined or calibrated special cases",
    "Supports effort/visit thresholds (VISITS and related downgrades)",
    "Flags index (partial-coverage) estimates (relative-abundance context)",
    "Legacy qualitative uncertainty field; not a substitute for quantified metadata",
    "Legacy qualitative uncertainty field; not a substitute for quantified metadata",
    "Legacy/import field (historical); not consistently present",
    "QA/workflow stage (preliminary/near final/final); not a type determinant"
  ),
  check.names = FALSE
)

if (file.exists(dd_path)) {
  dd <- read.csv(
    dd_path,
    stringsAsFactors = FALSE,
    check.names = FALSE,
    fileEncoding = "latin1"
  )

  dd_small <- dd[dd[["Field Name"]] %in% wanted_fields, c("Field Name"), drop = FALSE]
  dd_small <- merge(dd_small, field_roles, by = "Field Name", all.x = TRUE)
  dd_small <- dd_small[match(wanted_fields, dd_small[["Field Name"]]), ]
} else {
  dd_small <- data.frame(
    `Field Name` = wanted_fields,
    check.names = FALSE
  )
  dd_small <- merge(dd_small, field_roles, by = "Field Name", all.x = TRUE)
  dd_small <- dd_small[match(wanted_fields, dd_small[["Field Name"]]), ]
}

csas_kable(
  dd_small,
  escape = TRUE,
  bold_header = FALSE,
  font_size = 8,
  align = c("l", "l"),
  caption = "NuSEDS fields relevant to estimate-type classification and how they relate to the updated guidance (see the NuSEDS data dictionary in docs/context/)."
) |>
  csas_colspec(1, width = "3.5cm") |>
  csas_colspec(2, width = "10.5cm")
```
