`r if(knitr:::is_latex_output()) '% begin csasdown appendix'`
`r if(!knitr:::is_latex_output()) '# (APPENDIX) Appendix {-}'`

`r if (knitr::is_latex_output()) "\\\\clearpage"`

# Appendix A: Classification rule summary {#app:classification-rule-summary}

```{r classification-rule-summary-table}
classification_rule_summary <- data.frame(
  `Rule stage` = c(
    "1. Data format gate",
    "2. Method known gate",
    "3. Method family assignment",
    "4. Family-specific checks (counting systems)",
    "5. Family-specific checks (survey indices)",
    "6. Combined-method integration rule",
    "7. Documentation gate",
    "8. Precision/accuracy gate",
    "9. Final assignment"
  ),
  `Decision / trigger` = c(
    "Record is non-numeric (presence or not-detected)",
    "Enumeration and/or estimation method not identified",
    "Primary method family selected (FS, V, A, S, T, R, P, M)",
    "Coverage, uptime, bypass/breach, cross-section, and related counting-system checks fail or are incomplete",
    "Visit effort, timing relative to run, visibility, reach coverage, or related survey checks fail or are incomplete",
    "Multiple components contribute to final estimate",
    "Evidence package insufficient for independent interpretation",
    "Candidate Type requires uncertainty/accuracy evidence that is missing or weak",
    "All gates evaluated"
  ),
  `Classification effect` = c(
    "Assign Type 6",
    "Assign Provisional Type 5",
    "Select applicable rule path",
    "Apply conservative downgrades within selected path",
    "Apply conservative downgrades within selected path",
    "Final Type cannot exceed weakest component unless integration method supports stronger classification",
    "Apply conservative downgrade",
    "Apply conservative downgrade",
    "Return Final Type plus explicit qualifier codes"
  ),
  `Primary qualifier code(s)` = c(
    "(none)",
    "METHOD_UNKNOWN",
    "(path selection step)",
    "RUN_COVERAGE, UPTIME, BREACH_BYPASS, XSEC_COVERAGE, DEVICE_CONFIG",
    "VISITS, TIMING, VISIBILITY, REACH_COVERAGE, EFF_STD, DETECTABILITY",
    "INFILL_METHOD (as relevant), component-level qualifiers",
    "DOC",
    "PRECISION_ACCURACY",
    "All triggered qualifiers retained as audit trail"
  ),
  check.names = FALSE
)

csas_kable(
  classification_rule_summary,
  escape = TRUE,
  bold_header = FALSE,
  font_size = 8,
  align = c("l", "l", "l", "l"),
  caption = "Single-table summary of the NuSEDS estimate-type classification rules used for forward implementation."
) |>
  csas_colspec(1, width = "3.6cm") |>
  csas_colspec(2, width = "5.2cm") |>
  csas_colspec(3, width = "3.4cm") |>
  csas_colspec(4, width = "3.6cm")
```

`r if (knitr::is_latex_output()) "\\\\clearpage"`

# Appendix B: Downgrade criteria codes {#app:downgrade-codes}

The qualifier-code table is formatted for high-density reference use. In PDF output it is rendered on a landscape page to improve line wrapping and scanability; HTML and DOCX outputs retain standard page flow.

`r if (knitr::is_latex_output()) "\\\\begin{landscape}"`

```{r downgrade-codes}
downgrade_code_terms <- c(
  "RunCoverage",
  "NumberOfVisits",
  "BreachBypass",
  "Uptime",
  "ReviewQA",
  "Visibility",
  "ReachCoverage",
  "Timing",
  "EffortStandardization",
  "CrossSectionCoverage",
  "Classification",
  "TrapEfficiency",
  "Detectability",
  "DeviceConfiguration",
  "EnvironmentalConditions",
  "MarkRecaptureAssumptions",
  "InfillMethod",
  "Documentation",
  "PrecisionAccuracy",
  "MethodUnknown"
)

downgrade_codes <- data.frame(
  Code = c(
    "RUN_COVERAGE",
    "VISITS",
    "BREACH_BYPASS",
    "UPTIME",
    "REVIEW_QA",
    "VISIBILITY",
    "REACH_COVERAGE",
    "TIMING",
    "EFF_STD",
    "XSEC_COVERAGE",
    "CLASSIFICATION",
    "TRAP_EFF",
    "DETECTABILITY",
    "DEVICE_CONFIG",
    "ENV_COND",
    "MR_ASSUMP",
    "INFILL_METHOD",
    "DOC",
    "PRECISION_ACCURACY",
    "METHOD_UNKNOWN"
  ),
  Description = c(
    "Run window coverage gaps (e.g., <95% of season days).",
    "Too few site visits for the claimed precision tier.",
    "Bypass or breach risk at a counting site.",
    "Device outages or unobserved intervals (e.g., uptime <95%).",
    "Insufficient review of automated events.",
    "Poor visibility or lighting limits detection.",
    "Intended reaches or segments not surveyed.",
    "Visits not aligned with migration peak or plateau.",
    "Inconsistent effort standardization.",
    "Partial cross-section coverage.",
    "Species or target classification error.",
    "Unmeasured or uncertain trap efficiency.",
    "Redd or carcass detectability bias.",
    "Device settings not documented or unstable.",
    "Environmental conditions bias detection.",
    "Mark-recapture assumption violations.",
    "Interpolation or infilling used; defensibility must be documented.",
    "Documentation or evidence missing.",
    "Precision or accuracy weaker than candidate Type.",
    "Survey or analysis method not identified."
  ),
  `Ontology term` = paste0("gcdfo:", downgrade_code_terms),
  `Ontology IRI` = paste0("https://w3id.org/gcdfo/salmon#", downgrade_code_terms),
  check.names = FALSE
)

csas_kable(
  downgrade_codes,
  escape = TRUE,
  bold_header = FALSE,
  font_size = 8,
  align = c("l", "l", "l", "l"),
  caption = "Downgrade criteria codes used by the classification key, with linkages to published DFO Salmon Ontology terms."
) |>
  csas_colspec(1, width = "2.8cm") |>
  csas_colspec(2, width = "8.2cm") |>
  csas_colspec(3, width = "2.6cm") |>
  csas_colspec(4, width = "6.0cm")
```

`r if (knitr::is_latex_output()) "\\\\end{landscape}"`


`r if (knitr::is_latex_output()) "\\\\clearpage"`

# Appendix C: NuSEDS data dictionary crosswalk {#app:nuseds-crosswalk}

This appendix documents how the updated guidance relates to NuSEDS fields and controlled vocabularies (see `docs/context/Data_Dictionary_NuSEDS_EN.csv`).

## C.1 Enumeration method values (ENUMERATION_METHODS)

```{r nuseds-enum-crosswalk}
dd_path <- file.path("docs", "context", "Data_Dictionary_NuSEDS_EN.csv")
dd_available <- file.exists(dd_path)

if (!dd_available) {
  dd <- NULL
  knitr::asis_output(
    "NuSEDS data dictionary CSV not present in this repo. Add it at `docs/context/Data_Dictionary_NuSEDS_EN.csv` to generate the crosswalk tables in Appendix C."
  )
} else {
  dd <- read.csv(
    dd_path,
    stringsAsFactors = FALSE,
    check.names = FALSE,
    fileEncoding = "latin1"
  )

  enum_def <- dd[["Field Definition"]][dd[["Field Name"]] == "ENUMERATION_METHODS"]
  enum_vals <- trimws(strsplit(sub(".*Values are:", "", enum_def), ",")[[1]])
  enum_vals <- sub("\\.$", "", enum_vals)

enum_crosswalk <- data.frame(
  `NuSEDS ENUMERATION_METHODS value` = enum_vals,
  `Suggested method family` = NA_character_,
  `Ontology term` = NA_character_,
  `WIDOCO term page` = NA_character_,
  Notes = "",
  check.names = FALSE
)

enum_crosswalk[["Suggested method family"]][enum_crosswalk[[1]] %in% c(
  "Bank Walk", "Stream Walk", "Walk", "Boat", "Float", "Snorkel", "Strip Counts",
  "Spot Checks", "Dead Pitch", "Peak Live and Dead Count"
)] <- "V"
enum_crosswalk[["Suggested method family"]][enum_crosswalk[[1]] %in% c(
  "Fence", "Electronic Counters", "Enumeration by Hatchery", "Broodstock Removal"
)] <- "FS"
enum_crosswalk[["Suggested method family"]][enum_crosswalk[[1]] %in% c(
  "Fixed Wing Aircraft", "Helicopter"
)] <- "A"
enum_crosswalk[["Suggested method family"]][enum_crosswalk[[1]] %in% c(
  "Hydroacoustic Station"
)] <- "S"
enum_crosswalk[["Suggested method family"]][enum_crosswalk[[1]] %in% c(
  "Trap"
)] <- "T"
enum_crosswalk[["Suggested method family"]][enum_crosswalk[[1]] %in% c(
  "Redd Counts"
)] <- "R"
enum_crosswalk[["Suggested method family"]][enum_crosswalk[[1]] %in% c(
  "Electroshocking"
)] <- "P"
enum_crosswalk[["Suggested method family"]][enum_crosswalk[[1]] %in% c(
  "Tag Recovery"
)] <- "M"
enum_crosswalk[["Suggested method family"]][enum_crosswalk[[1]] %in% c(
  "Based on Angling Catch"
)] <- "P"
enum_crosswalk[["Suggested method family"]][enum_crosswalk[[1]] %in% c(
  "Biologist/Working Group", "Other"
)] <- "unknown"

enum_widoco_base <- "https://dfo-pacific-science.github.io/dfo-salmon-ontology/#"

enum_crosswalk[["Ontology term"]][enum_crosswalk[[1]] %in% c(
  "Bank Walk", "Stream Walk", "Walk", "Boat", "Float", "Strip Counts", "Spot Checks", "Dead Pitch", "Peak Live and Dead Count"
)] <- "gcdfo:VisualGroundCount"
enum_crosswalk[["Ontology term"]][enum_crosswalk[[1]] %in% c("Snorkel", "Snorkel Swim")] <- "gcdfo:VisualSnorkelCount"
enum_crosswalk[["Ontology term"]][enum_crosswalk[[1]] %in% c(
  "Fence", "Enumeration by Hatchery", "Broodstock Removal"
)] <- "gcdfo:FixedSiteCensusManual"
enum_crosswalk[["Ontology term"]][enum_crosswalk[[1]] %in% c("Electronic Counters")] <- "gcdfo:FixedSiteCensusElectronic"
enum_crosswalk[["Ontology term"]][enum_crosswalk[[1]] %in% c("Fixed Wing Aircraft", "Helicopter")] <- "gcdfo:AerialSurveyCount"
enum_crosswalk[["Ontology term"]][enum_crosswalk[[1]] %in% c("Hydroacoustic Station")] <- "gcdfo:HydroacousticSonarCount"
enum_crosswalk[["Ontology term"]][enum_crosswalk[[1]] %in% c("Trap")] <- "gcdfo:TrapCount"
enum_crosswalk[["Ontology term"]][enum_crosswalk[[1]] %in% c("Redd Counts")] <- "gcdfo:ReddCount"
enum_crosswalk[["Ontology term"]][enum_crosswalk[[1]] %in% c("Electroshocking")] <- "gcdfo:ElectrofishingCount"
enum_crosswalk[["Ontology term"]][enum_crosswalk[[1]] %in% c("Tag Recovery")] <- "gcdfo:MarkRecaptureFieldProgram"
enum_crosswalk[["Ontology term"]][enum_crosswalk[[1]] %in% c("Based on Angling Catch")] <- "gcdfo:EnumerationMethod"

has_enum_term <- !is.na(enum_crosswalk[["Ontology term"]])
enum_crosswalk[["WIDOCO term page"]][has_enum_term] <-
  paste0(enum_widoco_base, sub("^gcdfo:", "", enum_crosswalk[["Ontology term"]][has_enum_term]))

enum_crosswalk[["Notes"]][enum_crosswalk[[1]] == "Peak Live and Dead Count"] <-
  "Value is analysis-like; prefer capturing peak/cumulative variants under ESTIMATE_METHOD."
enum_crosswalk[["Notes"]][enum_crosswalk[[1]] == "Dead Pitch"] <-
  "Carcass-based visual surveys; often paired with peak/cumulative dead estimation methods."
enum_crosswalk[["Notes"]][enum_crosswalk[[1]] == "Trap"] <-
  "If trap is non-spanning or efficiency-corrected, use T; fully constraining traps may behave more like fixed-site counting."
enum_crosswalk[["Notes"]][enum_crosswalk[[1]] == "Based on Angling Catch"] <-
  "Catch-based index; treat as a CPUE-style index unless more detail is provided; ontology linkage is provisional at scheme level."
enum_crosswalk[["Notes"]][enum_crosswalk[[1]] == "Biologist/Working Group"] <-
  "Not a method; treat as method-unknown unless a specific field/analysis method is documented elsewhere."
enum_crosswalk[["Notes"]][enum_crosswalk[[1]] == "Other"] <-
  "Treat as method-unknown unless a specific field/analysis method is documented elsewhere."

csas_kable(
  enum_crosswalk,
  escape = TRUE,
  bold_header = FALSE,
  font_size = 8,
  align = c("l", "c", "l", "l", "l"),
  caption = "Suggested mapping from NuSEDS enumeration method values to the updated key method families (FS, V, A, S, T, R, P, M), with ontology term and WIDOCO linkages."
) |>
  csas_colspec(1, width = "3.7cm") |>
  csas_colspec(2, width = "1.2cm") |>
  csas_colspec(3, width = "2.8cm") |>
  csas_colspec(4, width = "4.0cm") |>
  csas_colspec(5, width = "4.0cm")
}
```


```{r nuseds-enum-sankey, fig.cap="Sankey-style summary of how legacy NuSEDS enumeration-method terms converge to canonical parent terms used by the updated guidance.", fig.height=5.4, fig.width=6.0}
if (exists("enum_crosswalk")) {
  sank <- enum_crosswalk[, c("NuSEDS ENUMERATION_METHODS value", "Ontology term")]
  names(sank) <- c("source", "target")
  sank$target[is.na(sank$target) | sank$target == ""] <- "(unmapped / method-unknown)"

  # Order to reduce line crossings: group sources by target.
  sank <- sank[order(sank$target, sank$source), ]

  plot_sankey_like <- function(df, main = NULL) {
    src <- df$source
    tgt <- df$target

    targets <- unique(tgt)
    sources <- src

    # y-positions
    y_src <- seq(0.95, 0.05, length.out = length(sources))
    y_tgt <- seq(0.95, 0.05, length.out = length(targets))

    x1 <- 0.26
    x2 <- 0.74

    par(mar = c(0.6, 5.2, 2.0, 5.2), xpd = NA)
    plot.new()
    plot.window(xlim = c(0, 1), ylim = c(0, 1))

    # Draw links
    for (i in seq_along(sources)) {
      y1 <- y_src[i]
      y2 <- y_tgt[match(tgt[i], targets)]
      xspline(
        x = c(x1, 0.45, 0.55, x2),
        y = c(y1, y1, y2, y2),
        shape = 1,
        border = grDevices::adjustcolor("#2C7FB8", alpha.f = 0.35),
        lwd = 1
      )
    }

    # Draw nodes + labels
    text(x1 - 0.02, y_src, sources, adj = c(1, 0.5), cex = 0.62, xpd = NA)
    text(x2 + 0.02, y_tgt, targets, adj = c(0, 0.5), cex = 0.68, xpd = NA)

    if (!is.null(main)) {
      title(main = main, cex.main = 0.95)
    }
  }

  plot_sankey_like(sank, main = "Enumeration method terms → canonical parent terms")
} else {
  plot.new()
  text(0.5, 0.5, "NuSEDS data dictionary CSV not available; sankey figure not generated.")
}
```

## C.2 Estimate method values (ESTIMATE_METHOD)

`r if (knitr::is_latex_output()) "\\\\begin{landscape}"`

```{r nuseds-estimate-method-crosswalk}
if (!exists("dd_available") || !isTRUE(dd_available)) {
  knitr::asis_output(
    "NuSEDS data dictionary CSV not present in this repo. Add it at `docs/context/Data_Dictionary_NuSEDS_EN.csv` to generate the crosswalk tables in Appendix C."
  )
} else {
est_def <- dd[["Field Definition"]][dd[["Field Name"]] == "ESTIMATE_METHOD"]
est_lines <- unlist(strsplit(est_def, "\\n"))
est_lines <- trimws(est_lines)
est_lines <- est_lines[est_lines != "" & !grepl("There are several standard methods", est_lines)]
est_names <- trimws(sub(" - .*?$", "", est_lines))

est_crosswalk <- data.frame(
  `NuSEDS ESTIMATE_METHOD value` = est_names,
  `Guidance interpretation` = NA_character_,
  `Suggested method family` = NA_character_,
  `Ontology term` = NA_character_,
  `WIDOCO term page` = NA_character_,
  Notes = "",
  check.names = FALSE
)

est_crosswalk[["Guidance interpretation"]][est_crosswalk[[1]] %in% c(
  "Fixed Site Census", "Resistivity Counter", "Video Counter"
)] <- "Fixed-station tally / counter mode (often stored as estimate method)"
est_crosswalk[["Suggested method family"]][est_crosswalk[[1]] %in% c(
  "Fixed Site Census", "Resistivity Counter", "Video Counter"
)] <- "FS"

est_crosswalk[["Guidance interpretation"]][est_crosswalk[[1]] %in% c(
  "Sonar-ARIS", "Sonar-DIDSON"
)] <- "Hydroacoustic modelling pipeline"
est_crosswalk[["Suggested method family"]][est_crosswalk[[1]] %in% c(
  "Sonar-ARIS", "Sonar-DIDSON"
)] <- "S"

est_crosswalk[["Guidance interpretation"]][est_crosswalk[[1]] %in% c(
  "Mark & Recapture: Petersen",
  "Mark & Recapture: Jolly-Seber",
  "Mark & Recapture: Bayesian",
  "Mark & Recapture: Open Model"
)] <- "Mark-recapture estimation"
est_crosswalk[["Suggested method family"]][est_crosswalk[[1]] %in% c(
  "Mark & Recapture: Petersen",
  "Mark & Recapture: Jolly-Seber",
  "Mark & Recapture: Bayesian",
  "Mark & Recapture: Open Model"
)] <- "M"

est_crosswalk[["Guidance interpretation"]][est_crosswalk[[1]] %in% c(
  "Area Under the Curve",
  "Peak Live + Dead",
  "Peak Live + Cumulative Dead"
)] <- "Survey-series estimation (AUC/peak variants; platform-specific ceiling depends on underlying survey family)"
est_crosswalk[["Suggested method family"]][est_crosswalk[[1]] %in% c(
  "Area Under the Curve",
  "Peak Live + Dead",
  "Peak Live + Cumulative Dead"
)] <- "depends"

est_crosswalk[["Guidance interpretation"]][est_crosswalk[[1]] %in% c(
  "(Peak Live+Cum Dead)*Expansion",
  "Peak Live * Expansion"
)] <- "Survey-series expansion (inherits the underlying survey family and any weaker component)"
est_crosswalk[["Suggested method family"]][est_crosswalk[[1]] %in% c(
  "(Peak Live+Cum Dead)*Expansion",
  "Peak Live * Expansion"
)] <- "depends"

est_crosswalk[["Guidance interpretation"]][est_crosswalk[[1]] %in% c("Redd Count")] <-
  "Redd-based estimation (default conservative survey/index pathway unless separately calibrated/modelled)"
est_crosswalk[["Suggested method family"]][est_crosswalk[[1]] %in% c("Redd Count")] <- "R"

est_crosswalk[["Guidance interpretation"]][est_crosswalk[[1]] %in% c("Cumulative CPUE")] <-
  "CPUE index"
est_crosswalk[["Suggested method family"]][est_crosswalk[[1]] %in% c("Cumulative CPUE")] <- "P"

est_crosswalk[["Guidance interpretation"]][est_crosswalk[[1]] %in% c(
  "Addition/Subtraction",
  "Multiplication/Division",
  "Lake Expansion"
)] <- "Math/expansion operations (depends on base method)"
est_crosswalk[["Suggested method family"]][est_crosswalk[[1]] %in% c(
  "Addition/Subtraction",
  "Multiplication/Division",
  "Lake Expansion"
)] <- "depends"

est_crosswalk[["Guidance interpretation"]][est_crosswalk[[1]] %in% c("Calibrated Time Series")] <-
  "Calibrated time series (requires calibration source + diagnostics)"
est_crosswalk[["Suggested method family"]][est_crosswalk[[1]] %in% c("Calibrated Time Series")] <- "depends"

est_crosswalk[["Guidance interpretation"]][est_crosswalk[[1]] %in% c("Combined Methods")] <-
  "Combined-method workflow (requires explicit component listing)"
est_crosswalk[["Suggested method family"]][est_crosswalk[[1]] %in% c("Combined Methods")] <- "depends"

est_crosswalk[["Guidance interpretation"]][est_crosswalk[[1]] %in% c(
  "Insufficient Information",
  "Unknown Estimate Method",
  "Other Estimate Method",
  "Not Applicable",
  "Expert Opinion",
  "Cumulative New"
)] <- "Method unknown/administrative label"
est_crosswalk[["Suggested method family"]][est_crosswalk[[1]] %in% c(
  "Insufficient Information",
  "Unknown Estimate Method",
  "Other Estimate Method",
  "Not Applicable",
  "Expert Opinion",
  "Cumulative New"
)] <- "unknown"

est_widoco_base <- "https://dfo-pacific-science.github.io/dfo-salmon-ontology/#"

est_crosswalk[["Ontology term"]][est_crosswalk[[1]] %in% c(
  "Fixed Site Census", "Resistivity Counter", "Video Counter"
)] <- "gcdfo:FixedStationTally"
est_crosswalk[["Ontology term"]][est_crosswalk[[1]] %in% c(
  "Sonar-ARIS", "Sonar-DIDSON"
)] <- "gcdfo:HydroacousticModelling"
est_crosswalk[["Ontology term"]][est_crosswalk[[1]] %in% c(
  "Mark & Recapture: Petersen",
  "Mark & Recapture: Jolly-Seber",
  "Mark & Recapture: Bayesian",
  "Mark & Recapture: Open Model"
)] <- "gcdfo:MarkRecaptureAnalysis"
est_crosswalk[["Ontology term"]][est_crosswalk[[1]] %in% c(
  "Area Under the Curve"
)] <- "gcdfo:AreaUnderTheCurve"
est_crosswalk[["Ontology term"]][est_crosswalk[[1]] %in% c(
  "Peak Live + Dead", "Peak Live + Cumulative Dead", "Cumulative New"
)] <- "gcdfo:PeakCountAnalysis"
est_crosswalk[["Ontology term"]][est_crosswalk[[1]] %in% c(
  "(Peak Live+Cum Dead)*Expansion", "Peak Live * Expansion", "Addition/Subtraction", "Multiplication/Division", "Lake Expansion"
)] <- "gcdfo:ExpansionMathematicalOperations"
est_crosswalk[["Ontology term"]][est_crosswalk[[1]] %in% c(
  "Redd Count"
)] <- "gcdfo:ReddExpansionAnalysis"
est_crosswalk[["Ontology term"]][est_crosswalk[[1]] %in% c(
  "Calibrated Time Series"
)] <- "gcdfo:CalibratedTimeSeries"
est_crosswalk[["Ontology term"]][est_crosswalk[[1]] %in% c(
  "Combined Methods", "Cumulative CPUE"
)] <- "gcdfo:EstimateMethod"

has_est_term <- !is.na(est_crosswalk[["Ontology term"]])
est_crosswalk[["WIDOCO term page"]][has_est_term] <-
  paste0(est_widoco_base, sub("^gcdfo:", "", est_crosswalk[["Ontology term"]][has_est_term]))

est_crosswalk[["Notes"]][est_crosswalk[[1]] == "Resistivity Counter"] <-
  "Treat as a fixed-station tally when the installation functions as the primary constrained-opening counter. Record species/event attribution, validation, detection-correction, and bypass/coverage/QA context in the supporting explanation fields; resistivity remains in the FS pathway under the current guidance."
est_crosswalk[["Notes"]][est_crosswalk[[1]] == "Video Counter"] <-
  "Treat as a fixed-station tally within FS; ensure QA review rate, attribution logic (if used), and uptime/coverage metadata are captured."
est_crosswalk[["Notes"]][est_crosswalk[[1]] == "Area Under the Curve"] <-
  "Treat AUC as an analysis label applied after survey-family selection: ground/snorkel series follow the V-family pathway, while aerial series remain capped by the A-family ceiling (Type 3) under the current conservative guidance."
est_crosswalk[["Notes"]][est_crosswalk[[1]] %in% c("Peak Live + Dead", "Peak Live + Cumulative Dead")] <-
  "Peak-count summaries inherit the underlying survey family. Aerial-derived peak summaries remain capped by the A-family ceiling unless a future explicit exception is adopted."
est_crosswalk[["Notes"]][est_crosswalk[[1]] %in% c("(Peak Live+Cum Dead)*Expansion", "Peak Live * Expansion")] <-
  "Expansion does not replace the underlying survey family; record the base method and any weaker component explicitly."
est_crosswalk[["Notes"]][est_crosswalk[[1]] == "Calibrated Time Series"] <-
  "For interpretability, record calibration source years, diagnostics, and revision history."
est_crosswalk[["Notes"]][est_crosswalk[[1]] == "Combined Methods"] <-
  "Decompose into components (e.g., sonar + visual apportionment; fence + visual during breach) and apply conservative classification."
est_crosswalk[["Notes"]][est_crosswalk[[1]] == "Redd Count"] <-
  "Default redd-count interpretation is conservatively capped at Type 3. Any higher-confidence redd-derived absolute-abundance estimate should be handled only as an explicit calibrated/model-based exception with supporting validation and documentation."
est_crosswalk[["Notes"]][est_crosswalk[[1]] == "Cumulative New"] <-
  "Legacy dictionary label with no method definition in the source description; mapped to PeakCountAnalysis pending local confirmation."
est_crosswalk[["Notes"]][est_crosswalk[[1]] == "Cumulative CPUE"] <-
  "No specific CPUE estimate concept is currently defined in this scheme; linked at EstimateMethod scheme level."

# Fail-safe: prevent blank crosswalk rows if new dictionary values appear.
unmapped_est <- is.na(est_crosswalk[["Guidance interpretation"]]) |
  is.na(est_crosswalk[["Suggested method family"]])
if (any(unmapped_est)) {
  est_crosswalk[["Guidance interpretation"]][unmapped_est] <-
    "Unmapped dictionary value (conservative handling)"
  est_crosswalk[["Suggested method family"]][unmapped_est] <- "unknown"
  est_crosswalk[["Notes"]][unmapped_est & est_crosswalk[["Notes"]] == ""] <-
    "New value not yet mapped in this report; treat as method unknown until mapping is reviewed."
}

csas_kable(
  est_crosswalk,
  escape = TRUE,
  bold_header = FALSE,
  font_size = 8,
  align = c("l", "l", "c", "l", "l", "l"),
  caption = "Suggested interpretation of NuSEDS estimate method values under the updated guidance, with ontology term and WIDOCO linkages. Where 'depends' appears, additional method metadata are required to apply the key cleanly."
) |>
  csas_colspec(1, width = "3.4cm") |>
  csas_colspec(2, width = "4.4cm") |>
  csas_colspec(3, width = "1.4cm") |>
  csas_colspec(4, width = "3.0cm") |>
  csas_colspec(5, width = "3.8cm") |>
  csas_colspec(6, width = "7.2cm")
}
```

`r if (knitr::is_latex_output()) "\\\\end{landscape}"`


```{r nuseds-estimate-sankey, fig.cap="Sankey-style summary of how legacy NuSEDS estimate-method terms converge to canonical parent terms used by the updated guidance.", fig.height=5.8, fig.width=6.0}
if (exists("est_crosswalk")) {
  sank <- est_crosswalk[, c("NuSEDS ESTIMATE_METHOD value", "Ontology term")]
  names(sank) <- c("source", "target")
  sank$target[is.na(sank$target) | sank$target == ""] <- "(unmapped / method-unknown)"

  sank <- sank[order(sank$target, sank$source), ]

  plot_sankey_like <- function(df, main = NULL) {
    src <- df$source
    tgt <- df$target

    targets <- unique(tgt)
    sources <- src

    y_src <- seq(0.95, 0.05, length.out = length(sources))
    y_tgt <- seq(0.95, 0.05, length.out = length(targets))

    x1 <- 0.26
    x2 <- 0.74

    par(mar = c(0.6, 5.2, 2.0, 5.2), xpd = NA)
    plot.new()
    plot.window(xlim = c(0, 1), ylim = c(0, 1))

    for (i in seq_along(sources)) {
      y1 <- y_src[i]
      y2 <- y_tgt[match(tgt[i], targets)]
      xspline(
        x = c(x1, 0.45, 0.55, x2),
        y = c(y1, y1, y2, y2),
        shape = 1,
        border = grDevices::adjustcolor("#2C7FB8", alpha.f = 0.35),
        lwd = 1
      )
    }

    text(x1 - 0.02, y_src, sources, adj = c(1, 0.5), cex = 0.60, xpd = NA)
    text(x2 + 0.02, y_tgt, targets, adj = c(0, 0.5), cex = 0.66, xpd = NA)

    if (!is.null(main)) {
      title(main = main, cex.main = 0.95)
    }
  }

  plot_sankey_like(sank, main = "Estimate method terms → canonical parent terms")
} else {
  plot.new()
  text(0.5, 0.5, "NuSEDS data dictionary CSV not available; sankey figure not generated.")
}
```

## C.3 Type-to-accuracy/precision lookup used in the Escapement Estimate Type Toolkit

The Escapement Estimate Type Toolkit includes a compact lookup that can be used when populating legacy-style accuracy/precision class fields alongside estimate Type. This is a compatibility aid for workflows that still carry those legacy labels; it does **not** change the Type assignment itself.

```{r type-accuracy-precision-lookup}
type_lookup <- data.frame(
  `Estimate Type` = c(1, 2, 3, 4, 5, 6),
  `accuracy_class` = c(
    "actual_very_high",
    "actual_or_assigned_high",
    "assigned_range_med_high",
    "unknown_assumed_constant",
    "unknown_assumed_highly_variable",
    "medium_to_high"
  ),
  `precision_class` = c(
    "plus_minus_zero",
    "high_to_moderate",
    "assigned_med_high",
    "unknown_assumed_constant",
    "unknown_assumed_highly_variable",
    "unknown"
  ),
  check.names = FALSE
)

csas_kable(
  type_lookup,
  escape = TRUE,
  bold_header = FALSE,
  font_size = 8,
  align = c("c", "l", "l"),
  caption = "Escapement Estimate Type Toolkit lookup from estimate Type to legacy-style accuracy/precision class labels, included as a compatibility aid for workflows that still populate those fields."
) |>
  csas_colspec(1, width = "2.0cm") |>
  csas_colspec(2, width = "6.5cm") |>
  csas_colspec(3, width = "6.5cm")
```

# Appendix D: NuSEDS method-term summary provenance (SEN exploratory) {#app:nuseds-method-summary}

This appendix documents the exploratory method-vocabulary evidence used to build the crosswalk logic in Appendix C and to interpret the frequency snapshot shown in Figure 2 of the main text.

## D.1 Source and processing notes

The method-term review is intended as implementation context, not as a frozen inventory of future NuSEDS values. The same source and processing notes applied to the Figure 2 frequency summary and to the Appendix C crosswalk review.

```{r method-summary-provenance}
method_summary_provenance <- data.frame(
  `Item` = c(
    "Source extract",
    "Fields reviewed",
    "Year filter",
    "Term handling",
    "Primary outputs"
  ),
  `Details` = c(
    "All-area SEN method extract assembled for this report.",
    "`ENUMERATION_METHODS` and `ESTIMATE_METHOD`.",
    "Records with `ANALYSIS_YR` between 1990 and 2025.",
    "Multi-valued cells split on semicolons, commas, and slashes; whitespace normalized before counting.",
    "Figure 2 in the main text (top terms) and the interpretive crosswalk context summarized in Appendix C."
  ),
  check.names = FALSE
)

csas_kable(
  method_summary_provenance,
  escape = TRUE,
  bold_header = FALSE,
  font_size = 8,
  align = c("l", "l"),
  caption = "Source and processing notes for the exploratory NuSEDS method-term review used in Figure 2 and Appendix C."
) |>
  csas_colspec(1, width = "3.2cm") |>
  csas_colspec(2, width = "11.0cm")
```

## D.2 Interpretation note

Because these summaries are derived from an exploratory extract rather than from a versioned public release, they should be interpreted as a dated snapshot that helps explain the crosswalk design---not as a guarantee that the same term frequencies will persist as NuSEDS continues to evolve.

`r if(knitr:::is_latex_output()) '% end csasdown appendix'`
